A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-05-28 paper/paper.tex
paper/contribution: polish

\subsection*{An iterated parametricity-based indexed construction}
The main contribution of the paper is to describe in full details a recipe uniformly characterizing unary and binary iterated parametricity in indexed form and to derive from it a new indexed presentation, called indexed \emph{$\nu$-sets}, uniformly of augmented semi-simplicial and semi-cubical sets. Alternatively, from a category-theory perspective, we are characterizing, among the class of matching objects up to isomorphism for augmented semi-simplicial and cubical sets, a particular instance that reflects the structure of iterated parametricity.

\subsection*{Contribution}
Our work:
\begin{itemize}
  \item Characterizes, will full details, how to technically define parametricity iterated in arbitrary dimensions.
  \item Provides a new example of indexed constructions of semi-simplicial and semi-cubical sets~\citep{voevodsky12,herbelin15,part15}, useful in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\citep{part15,shulman14,altenkirch16,kraus17}, as well as the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett}).
  \item Makes a step in the direction of the program initiated in \cite{altenkirch15} to develop parametricity-based models of parametric type theory~\citep{bernardy15,nuyts17,cavallo19} and cubical type theory~\citep{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do; compare, for example, the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~(\citeyear{curien14}).
\end{itemize}
\subsection*{Outline}
The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we give intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.

>>>>

\subsection*{Contribution}
The main contribution of the paper is to describe in full details a recipe uniformly characterizing unary and binary iterated parametricity in indexed form and to derive from it a new indexed presentation, called indexed \emph{$\nu$-sets}, uniformly of augmented semi-simplicial and semi-cubical sets. Alternatively, from a category-theory perspective, we are characterizing, among the class of matching objects up to isomorphism for augmented semi-simplicial and cubical sets, a particular instance that reflects the structure of iterated parametricity. Additionally,

\begin{itemize}
  \item This new construction is useful in the direction of better understanding the technicality of coherence issues in defining ``semi-simplicial types''~\citep{part15,shulman14,altenkirch16,kraus17}, as well as the dialectic between reasoning in Extensional Type Theory and Intensional Type Theory (as will be discussed in Section~\ref{sec:ett}).
  \item Our work is a step in the direction of the program initiated in \cite{altenkirch15} to develop parametricity-based models of parametric type theory~\citep{bernardy15,nuyts17,cavallo19} and cubical type theory~\citep{bezem13,cohen16,angiuli21} which are closer to the syntax of type theory and thus more liable to reflect definitional properties of type theory than presheaf-based cubical sets would do; compare, for example, the loss of definitional properties when interpreting ``indexed'' dependent types of type theory as ``fibrations'' in models such as locally cartesian closed categories~\citep{curien14}.
\end{itemize}

\subsection*{Outline}
The outline of the paper is as follows. We recall the definition of the augmented semi-simplicial and semi-cubical categories using a combinatorial presentation, that directly leads to our generalization to $\nu$-sets, in \ref{sec:nu}. Before spelling out the details of our formalization in type-theoretic language in \ref{sec:ett} and \ref{sec:itt}, we give intuitions in \ref{sec:intuition}. We end with some of the finer details of our mechanization in \ref{sec:mech}.
